---
title: "Data Import & Quality Assurance"
subtitle: "A Parallel Guide in R and Python"
---


# Import and Merge Data

As mentioned before, we will be using the data from this study by Santos & Duarte (2016): 
https://peerj.com/articles/2648/ . If you want to run the code below on your local machine, please follow the links in their PeerJ article to download their data from FigShare. Alternatively, if you just want to follow along with the code, you can see all of the code, outputs, and figures below without having to download any data. 

If you do download the data, please save it in this structure:

```text
YOUR_FOLDER_HERE/               # Name of the project folder on your local computer
└── data/                       # data subfolder
    └── BDS/                
        ├── trials/              # These are the individual trials data files downloaded from Figshare
            └── BDS00001.csv         # 60s of stabilometer data for trial 1
            └── ...
            └── BDS01956.csv         # 60s of stabilometer data for trial 1956
        └── output/             # This is where we will store outputs generated by the code
        └── BDSinfo.csv         # This is the demographic datafile from FigShare. This file maps trials to subjects.
        # Note that on figshare, the BDSinfo.csv is in the same folder as the individual trial data
        # we move it out of that folder to make some of our automated steps easier.
        
        
```
**Note that our data differ from the original Santos & Duarte (2016) dataset.** The original data were collected and processed to a very high standard to maximize their utility for biomechanics researchers. Because we want to use these data for teaching purposes, however, we manipulated some of the original data to introduce artifacts and other types of errors that we want to catch in our pipeline. Thus, the data presented below are for teaching purposes only and we explicitly do not make the raw data available to avoid confusion. If readers are interested in following along step by step, they should download the original data from their source, but remember that those data will not include the errors we built in for teaching purposes.


In the code chunks below, you can toggle between the `R` and `Python` implementations. We have tried to keep these code as similar as possible between the two languages, but it is important to remember that:

1. There is never only one way to things in programming, although some ways are better than others (and we try to show only "good" ways).

2. Each programming language is different. So, even though we try to keep the code as structurally similar as possible, there often won't be a one-to-one translation between languages. 

In the chunks below, we show the structure of the "raw" data after they have been merged into one master data files (rather than individual .csv files for every subject on every trial). 

::: panel-tabset
## R (`tidyverse` & `data.table`)

```{r}
#| label: script00-r
library(reticulate)
use_condaenv("r-reticulate", required = TRUE)

library(tidyverse)
library(data.table)
library(arrow)

# 1. Read metadata
info_path <- "./data/BDS/BDSinfo.txt"
info_raw  <- read_tsv(info_path, show_col_types = FALSE)

info <- info_raw %>%
  select(Trial, Subject, Vision, Surface, Age, AgeGroup, Gender, Height, Weight, BMI) %>%
  mutate(
    Subject  = as.integer(Subject),
    Vision   = factor(Vision, levels = c("Open", "Closed")),
    Surface  = factor(Surface, levels = c("Firm", "Foam")),
    AgeGroup = factor(AgeGroup),
    Gender   = factor(Gender)
  )

# 2. Batch-read raw trial files
trial_files <- list.files("./data/BDS/trials", pattern = "^BDS\\d+\\.txt$", full.names = TRUE)
trial_ids   <- tools::file_path_sans_ext(basename(trial_files))

read_one_trial <- function(path, trial_id) {
  fread(path, header = TRUE, sep = "\t", showProgress = FALSE) %>%
    as_tibble() %>%
    rename_with(~ str_trim(str_remove(.x, "\\[.*\\]"))) %>%
    mutate(Trial = trial_id, .before = 1)
}

trial_list <- map2(trial_files, trial_ids, read_one_trial)
all_trials <- bind_rows(trial_list)

# 3. Merge & Save
master <- all_trials %>%
  left_join(info, by = "Trial") %>%
  relocate(Trial, Subject, Vision, Surface, Age, AgeGroup, Gender, Height, Weight, BMI, .before = Time)

dir.create("./data/BDS/output", recursive = TRUE, showWarnings = FALSE)
arrow::write_parquet(master, "./data/BDS/output/BDS_master.parquet")
glimpse(master)
```

## Python (`pandas` & `pyarrow`)

```{python}
#| label: script00-python
import glob
import os
import pandas as pd

# 1. Read metadata
info_raw = pd.read_csv("./data/BDS/BDSinfo.txt", sep="\t")
cols = ["Trial", "Subject", "Vision", "Surface", "Age", "AgeGroup", "Gender", "Height", "Weight", "BMI"]

info = info_raw[cols].copy()
info["Subject"] = info["Subject"].astype(int)
info["Vision"]  = pd.Categorical(info["Vision"], categories=["Open", "Closed"])
info["Surface"] = pd.Categorical(info["Surface"], categories=["Firm", "Foam"])

# 2. Batch-read raw trial files
trial_files = sorted(glob.glob("./data/BDS/trials/BDS*.txt"))

def read_one_trial(path):
    trial_id = os.path.splitext(os.path.basename(path))[0]
    df = pd.read_csv(path, sep="\t")
    # Clean headers (strip units like [cm], [N])
    df.columns = df.columns.str.replace(r"\[.*\]", "", regex=True).str.strip()
    df.insert(0, "Trial", trial_id)
    return df

trial_list = [read_one_trial(f) for f in trial_files]
all_trials = pd.concat(trial_list, ignore_index=True)

# 3. Merge & Save
master = pd.merge(all_trials, info, on="Trial", how="left")
os.makedirs("./data/BDS/output", exist_ok=True)
master.to_parquet("./data/BDS/output/BDS_master.parquet")
master.info()
```
:::

The first few columns track **structural** information about how these data are 
organized:

* `Trial` - the trial number which was used to index the original raw data files.
* `Subject` - the unique subject identification number, each subject has ~4 trials and ~60s worth of data per trial.
* `Vision` - was the trial completed with eyes open or eyes closed.
* `Surface` - was the trial completed on a firm surface or a foam surface.
* `Time` - time, in seconds, within each trial. Trials were, in general, 60s long.     

... and the last few columns are the key dependent variables tracked continuously throughout each trial:

* `Fx`, `Fy`, and `Fz` - force in each plane of the force plate. 
* `Mx`, `My`, and `Mz` - moment arms corresponding to the forces in each plane.
* `CoPx` and `CoPy` - the center of pressure as a point, in the x- and y-planes of force plate, calculated from the preceding forces and moment arms. 



------------------------------------------------------------------------

# Quality Assurance & Feature Extraction

From raw Center of Pressure (COP) time-series data $(COP_x, COP_y)$, we will calculate key biomechanical variables that a researcher might be interested in:

1. **Path Lengths**: Sum of absolute step distances across axes:
   $$\text{PathLength}_{\text{AP}} = \sum_{n=1}^{N-1} |COP_{y, n+1} - COP_{y, n}|$$
   $$\text{PathLength}_{\text{Total}} = \sum_{n=1}^{N-1} \sqrt{(COP_{x, n+1} - COP_{x, n})^2 + (COP_{y, n+1} - COP_{y, n})^2}$$

2. **95% Confidence Ellipse Area & Boundary**:
   $$\text{Area}_{95} = 2 \pi F_{0.95, 2, N-2} \sqrt{\lambda_1 \lambda_2}$$
   where $\lambda_1, \lambda_2$ are eigenvalues of the $(COP_x, COP_y)$ covariance matrix.

::: {.panel-tabset}
## R (`tidyverse` & `patchwork`)

```{r}
#| label: script01-r
#| fig-width: 9
#| fig-height: 8

library(tidyverse)
library(arrow)
library(patchwork)

master <- arrow::read_parquet("./data/BDS/output/BDS_master.parquet")

# 1. Feature Extraction & Ellipse Math
compute_path_metrics <- function(x, y, duration_s) {
  dx <- diff(x)
  dy <- diff(y)
  pl_ml    <- sum(abs(dx))
  pl_ap    <- sum(abs(dy))
  pl_total <- sum(sqrt(dx^2 + dy^2))
  
  tibble(
    PathLength_AP    = pl_ap,
    PathLength_ML    = pl_ml,
    PathLength_Total = pl_total,
    MeanVelocity_AP  = pl_ap / duration_s,
    MeanVelocity_ML  = pl_ml / duration_s
  )
}

compute_ellipse95 <- function(x, y) {
  n <- length(x)
  cov_mat <- stats::cov(cbind(x, y))
  eig <- eigen(cov_mat)
  Fval <- stats::qf(0.95, df1 = 2, df2 = n - 2)
  area <- 2 * pi * Fval * sqrt(eig$values[1] * eig$values[2])
  
  list(area = area, eig = eig, Fval = Fval, center = c(mean(x), mean(y)))
}

ellipse_boundary_df <- function(x, y, n_points = 100) {
  e <- compute_ellipse95(x, y)
  axis_lengths <- sqrt(2 * e$Fval * e$eig$values)
  theta  <- seq(0, 2 * pi, length.out = n_points)
  circle <- rbind(cos(theta), sin(theta))
  pts <- e$eig$vectors %*% diag(axis_lengths) %*% circle
  
  tibble(ex = e$center[1] + pts[1, ], ey = e$center[2] + pts[2, ])
}

compute_trial_features <- function(df) {
  n        <- nrow(df)
  duration <- max(df$Time) - min(df$Time)
  path_m   <- compute_path_metrics(df$COPx, df$COPy, duration)
  ellipse  <- compute_ellipse95(df$COPx, df$COPy)
  time_d   <- diff(df$Time)
  
  bind_cols(
    path_m,
    tibble(
      EllipseArea95     = ellipse$area,
      n_samples         = n,
      duration_s        = duration,
      Fz_min            = min(df$Fz),
      Fz_median         = median(df$Fz),
      time_gap_max      = max(time_d),
      time_nonmonotonic = any(time_d <= 0)
    )
  )
}

# Process trial summary
trial_features <- master %>%
  group_by(Trial, Subject, Vision, Surface, Age, AgeGroup, Gender, Height, Weight, BMI) %>%
  group_modify(~ compute_trial_features(.x)) %>%
  ungroup()

# 2. Structural & Signal QA Checks
modal_n     <- as.integer(names(sort(table(trial_features$n_samples), decreasing = TRUE))[1])
expected_dt <- 0.01

trial_features <- trial_features %>%
  mutate(
    flag_sample_count   = abs(n_samples - modal_n) > 10,
    flag_time_gap       = time_gap_max > (expected_dt * 2) | time_nonmonotonic,
    flag_fz_unweighting = Fz_min < (0.7 * Fz_median),
    flag_any            = flag_sample_count | flag_time_gap | flag_fz_unweighting
  )

# Save QA Logs & Summaries
qa_log <- trial_features %>% filter(flag_any)
write_csv(qa_log, "./data/BDS/output/qa_log.csv")
saveRDS(trial_features, "./data/BDS/output/BDS_trial_summary.rds")
write_csv(trial_features, "./data/BDS/output/BDS_trial_summary.csv")

# 3. Overview Outlier-Screening Figure
overview_data <- trial_features %>%
  select(Trial, Subject, Vision, Surface, PathLength_Total, MeanVelocity_AP, MeanVelocity_ML, EllipseArea95) %>%
  pivot_longer(cols = c(PathLength_Total, MeanVelocity_AP, MeanVelocity_ML, EllipseArea95),
               names_to = "Measure", values_to = "Value")

p_overview <- ggplot(overview_data, aes(x = Surface, y = Value, fill = Vision)) +
  geom_boxplot(outlier.alpha = 0.5, position = position_dodge(width = 0.8)) +
  facet_wrap(~ Measure, scales = "free_y") +
  theme_minimal(base_size = 11) +
  labs(title = "Trial-level DV Distributions (Quality Assurance Overview)", y = NULL)

print(p_overview)

# 4. Per-Subject Diagnostic Figures (Cumulative Path + 95% Ellipses)
master_aug <- master %>%
  arrange(Trial, Time) %>%
  group_by(Trial) %>%
  mutate(
    StepDist       = sqrt(c(0, diff(COPx))^2 + c(0, diff(COPy))^2),
    CumulativePath = cumsum(StepDist)
  ) %>%
  ungroup()

rep_lookup <- master_aug %>%
  distinct(Trial, Subject, Vision, Surface) %>%
  arrange(Subject, Vision, Surface, Trial) %>%
  group_by(Subject, Vision, Surface) %>%
  mutate(Rep = factor(row_number())) %>%
  ungroup() %>%
  select(Trial, Rep)

master_aug <- master_aug %>% left_join(rep_lookup, by = "Trial")

make_subject_qa_plot <- function(subject_id, df) {
  p_top <- ggplot(df, aes(x = Time, y = CumulativePath, group = Trial, color = Rep, linetype = Vision)) +
    geom_line(linewidth = 0.4) +
    facet_wrap(~ Surface) +
    labs(title = paste("Subject", subject_id, "- QA Diagnostic"),
         subtitle = "Cumulative COP Path Length (slope ~ velocity; jumps -> possible artifact)",
         x = "Time (s)", y = "Cumulative path (cm)") +
    theme_minimal(base_size = 9)
  
  ellipses <- df %>%
    group_by(Trial, Vision, Surface, Rep) %>%
    group_modify(~ ellipse_boundary_df(.x$COPx, .x$COPy)) %>%
    ungroup()
  
  p_bottom <- ggplot(df, aes(x = COPx, y = COPy, group = Trial, color = Rep, linetype = Vision)) +
    geom_path(linewidth = 0.3, alpha = 0.5) +
    geom_path(data = ellipses, aes(x = ex, y = ey, group = Trial), linewidth = 0.5, alpha = 0.9) +
    facet_wrap(~ Surface) +
    coord_equal() +
    labs(subtitle = "Stabilogram with 95% Confidence Ellipses", x = "COPx (cm)", y = "COPy (cm)") +
    theme_minimal(base_size = 9)
  
  (p_top / p_bottom) + plot_layout(guides = "collect")
}

# Save per-subject plots to directory and display Subject 1 as example
qa_dir <- "./data/BDS/output/qa_plots"
dir.create(qa_dir, recursive = TRUE, showWarnings = FALSE)

subject_ids <- unique(master_aug$Subject)
for (sid in subject_ids) {
  df_sub <- master_aug %>% filter(Subject == sid)
  p <- make_subject_qa_plot(sid, df_sub)
  ggsave(file.path(qa_dir, sprintf("subject_%03d_QA.png", sid)), p, width = 8, height = 9, dpi = 100)
}

# Display Subject 1 in document
make_subject_qa_plot(subject_ids[1], master_aug %>% filter(Subject == subject_ids[1]))
```

## Python (`numpy`, `scipy` & `matplotlib`)

```{python}
#| label: script01-python
#| fig-width: 9
#| fig-height: 8

import os
import numpy as np
import pandas as pd
from scipy.stats import f
import matplotlib.pyplot as plt
import seaborn as sns

master = pd.read_parquet("./data/BDS/output/BDS_master.parquet")

# 1. Feature Extraction & Ellipse Math
def compute_ellipse95(x, y):
    n = len(x)
    cov_mat = np.cov(x, y)
    eigvals, eigvecs = np.linalg.eigh(cov_mat)
    f_val = f.ppf(0.95, 2, n - 2)
    area = 2 * np.pi * f_val * np.sqrt(np.maximum(0, eigvals[0] * eigvals[1]))
    return area, eigvals, eigvecs, f_val, np.mean(x), np.mean(y)

def ellipse_boundary_df(x, y, n_points=100):
    area, eigvals, eigvecs, f_val, cx, cy = compute_ellipse95(x, y)
    axis_lengths = np.sqrt(2 * f_val * np.maximum(0, eigvals))
    theta = np.linspace(0, 2 * np.pi, n_points)
    circle = np.vstack([np.cos(theta), np.sin(theta)])
    pts = eigvecs @ np.diag(axis_lengths) @ circle
    return pd.DataFrame({'ex': cx + pts[0, :], 'ey': cy + pts[1, :]})

def compute_trial_features(df):
    duration = df["Time"].max() - df["Time"].min()
    dx = np.diff(df["COPx"])
    dy = np.diff(df["COPy"])
    
    pl_ml  = np.sum(np.abs(dx))
    pl_ap  = np.sum(np.abs(dy))
    pl_tot = np.sum(np.sqrt(dx**2 + dy**2))
    
    area, _, _, _, _, _ = compute_ellipse95(df["COPx"], df["COPy"])
    time_diffs = np.diff(df["Time"])
    
    return pd.Series({
        "PathLength_AP": pl_ap, "PathLength_ML": pl_ml, "PathLength_Total": pl_tot,
        "MeanVelocity_AP": pl_ap / duration, "MeanVelocity_ML": pl_ml / duration,
        "EllipseArea95": area, "n_samples": len(df), "duration_s": duration,
        "Fz_min": df["Fz"].min(), "Fz_median": df["Fz"].median(),
        "time_gap_max": time_diffs.max(), "time_nonmonotonic": (time_diffs <= 0).any()
    })

meta_cols = ["Trial", "Subject", "Vision", "Surface", "Age", "AgeGroup", "Gender", "Height", "Weight", "BMI"]
trial_summary = master.groupby(meta_cols, observed=True).apply(compute_trial_features).reset_index()

# 2. QA Flags
modal_n = trial_summary["n_samples"].mode()[0]
trial_summary["flag_sample_count"]   = (trial_summary["n_samples"] - modal_n).abs() > 10
trial_summary["flag_time_gap"]       = (trial_summary["time_gap_max"] > 0.02) | trial_summary["time_nonmonotonic"]
trial_summary["flag_fz_unweighting"] = trial_summary["Fz_min"] < (0.7 * trial_summary["Fz_median"])
trial_summary["flag_any"]            = trial_summary["flag_sample_count"] | trial_summary["flag_time_gap"] | trial_summary["flag_fz_unweighting"]

qa_log = trial_summary[trial_summary["flag_any"]]
qa_log.to_csv("./data/BDS/output/qa_log.csv", index=False)
trial_summary.to_csv("./data/BDS/output/BDS_trial_summary.csv", index=False)

# 3. Overview Outlier-Screening Figure
overview_data = trial_summary.melt(
    id_vars=["Trial", "Subject", "Vision", "Surface"],
    value_vars=["PathLength_Total", "MeanVelocity_AP", "MeanVelocity_ML", "EllipseArea95"],
    var_name="Measure", value_name="Value"
)

g = sns.catplot(
    data=overview_data, x="Surface", y="Value", hue="Vision", col="Measure",
    col_wrap=2, kind="box", height=3.5, aspect=1.2, sharey=False
)
g.fig.subplots_adjust(top=0.9)
g.fig.suptitle("Trial-level DV Distributions (Quality Assurance Overview)")
plt.show()

# 4. Per-Subject Diagnostic Figures with Faceting by Surface
master_aug = master.sort_values(["Trial", "Time"]).copy()
master_aug["dx"] = master_aug.groupby("Trial")["COPx"].diff().fillna(0)
master_aug["dy"] = master_aug.groupby("Trial")["COPy"].diff().fillna(0)
master_aug["StepDist"] = np.sqrt(master_aug["dx"]**2 + master_aug["dy"]**2)
master_aug["CumulativePath"] = master_aug.groupby("Trial")["StepDist"].cumsum()

# Add Rep indicator
rep_df = master_aug[["Trial", "Subject", "Vision", "Surface"]].drop_duplicates().sort_values(["Subject", "Vision", "Surface", "Trial"])
rep_df["Rep"] = rep_df.groupby(["Subject", "Vision", "Surface"]).cumcount() + 1
master_aug = master_aug.merge(rep_df[["Trial", "Rep"]], on="Trial")

def make_subject_qa_plot(subject_id, df_sub):
    fig, axes = plt.subplots(2, 2, figsize=(9, 8), sharex='row', sharey='row')
    palette = {1: "#1E6654", 2: "#CC5500", 3: "#5B7C99"}
    surfaces = ["Firm", "Foam"]
    
    for j, surf in enumerate(surfaces):
        df_surf = df_sub[df_sub["Surface"] == surf]
        ax_top = axes[0, j]
        ax_bot = axes[1, j]
        
        # Top Panel: Cumulative Path (Faceted by Surface)
        for (tr, rep, vis), df_tr in df_surf.groupby(["Trial", "Rep", "Vision"], observed=True):
            ls = "-" if vis == "Open" else "--"
            c = palette[rep]
            ax_top.plot(df_tr["Time"], df_tr["CumulativePath"], linestyle=ls, color=c, alpha=0.8)
        
        ax_top.set_title(f"Surface: {surf}")
        ax_top.set_xlabel("Time (s)")
        if j == 0:
            ax_top.set_ylabel("Cumulative path (cm)")
            
        # Bottom Panel: Stabilogram + 95% Ellipses (Faceted by Surface)
        for (tr, rep, vis), df_tr in df_surf.groupby(["Trial", "Rep", "Vision"], observed=True):
            ls = "-" if vis == "Open" else "--"
            c = palette[rep]
            ax_bot.plot(df_tr["COPx"], df_tr["COPy"], linestyle=ls, color=c, alpha=0.3, linewidth=0.8)
            
            ell_df = ellipse_boundary_df(df_tr["COPx"], df_tr["COPy"])
            ax_bot.plot(ell_df["ex"], ell_df["ey"], linestyle=ls, color=c, alpha=0.9, linewidth=1.2)
            
        ax_bot.set_title(f"Surface: {surf}")
        ax_bot.set_xlabel("COPx (cm)")
        if j == 0:
            ax_bot.set_ylabel("COPy (cm)")
        # Fixed: Changed adjustable='datalim' to 'box' to support shared axes
        ax_bot.set_aspect("equal", adjustable="box")
    
    fig.suptitle(f"Subject {subject_id} - QA Diagnostic\nTop: Cumulative Path | Bottom: Stabilogram & 95% Ellipses", fontsize=12, fontweight="bold")
    plt.tight_layout()
    return fig

# Save all per-subject diagnostic PNGs
os.makedirs("./data/BDS/output/qa_plots", exist_ok=True)
subject_ids = master_aug["Subject"].unique()

for sid in subject_ids:
    df_sub = master_aug[master_aug["Subject"] == sid]
    fig = make_subject_qa_plot(sid, df_sub)
    fig.savefig(f"./data/BDS/output/qa_plots/subject_{sid:03d}_QA.png", dpi=100)
    if sid != subject_ids[0]:
        plt.close(fig)

# Display Subject 1 in Quarto output
plt.show()
```
:::

