---
title: "Static Data Visualization"
subtitle: "A Parallel Guide in R and Python"
---

# Static Data Visualization

In this section, we walk through the iterative process of building data visualizations. We start with raw observations, layer on aesthetic mappings and summary geometry, aggregate data to account for repeated measures, and conclude with multi-panel publication figures.

---

## Step 1: Raw Geometry (The Foundation)

We begin by mapping `Surface` condition to the x-axis and `PathLength_Total` to the y-axis. This raw point plot reveals the basic distribution of data but suffers from heavy overplotting across conditions.

It is worth noting that `plotnine` is a package for Python that uses syntax essentially identical to `ggplot` in R (both `plotnine` and `ggplot` are based on Leland Wilkinson’s "Grammar of Graphics" - a layered system to build data visualizations). However, when making this document using *Quarto* inside of *R Studio*, I need to call Python functions using `reticulate` and - for computing reasons I do not full understand - it is difficult to render `plotnine` objects through `reticulate`. Thus, for these examples, I am using `matplotlib` and `seaborn` to great plots similar to my `ggplot` visualizations, but for pragmatic users, it is probably more efficient to learn `ggplot`/`plotnine` syntax that is portable across languages.

::: {.panel-tabset}
## R (`ggplot2`)

```{r}
#| label: script02-r-plot1
#| fig-width: 6
#| fig-height: 4

library(tidyverse)
library(patchwork)
library(GGally)

DATA <- readRDS("./data/BDS/output/BDS_trial_summary.rds") %>% 
  mutate(Subject = factor(Subject),
         AgeGroup = fct_relevel(AgeGroup, "Young", "Old"))

# Plot 1: Basic scatter plot
ggplot(data = DATA, aes(x = Surface, y = PathLength_Total)) +
  geom_point()
```

## Python (`matplotlib` & `seaborn`)
```{python}
#| label: script02-py-plot1
#| fig-width: 6
#| fig-height: 4

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns

DATA = pd.read_csv("./data/BDS/output/BDS_trial_summary.csv")
DATA["AgeGroup"] = pd.Categorical(DATA["AgeGroup"], categories=["Young", "Old"], ordered=True)

# Step 1: Basic scatter plot
fig, ax = plt.subplots(figsize=(6, 4))
sns.scatterplot(data=DATA, x="Surface", y="PathLength_Total", color="black", ax=ax)
ax.set_ylabel("Total Path Length (cm)")
sns.despine()
plt.show()
```
:::

---

## Step 2: Adding Aesthetic Mappings

To evaluate the effect of visual input, we map `Vision` to point shape and color while dodging the points along the x-axis to reduce overlap.

::: {.panel-tabset}
## R (`ggplot2`)

```{r}
#| label: script02-r-plot2
#| fig-width: 6
#| fig-height: 4

# Plot 2: Dodged points by Vision condition
ggplot(data = DATA, aes(x = Surface, y = PathLength_Total)) +
  geom_point(aes(shape = Vision, col = Vision),
             position = position_dodge(width = 0.5)) +
  theme_bw()
```

## Python (`matplotlib` & `seaborn`)

```{python}
#| label: script02-py-plot2
#| fig-width: 6
#| fig-height: 4

import matplotlib.pyplot as plt
import seaborn as sns

# Step 2: Dodged points by Vision condition
fig, ax = plt.subplots(figsize=(6, 4))
sns.stripplot(
    data=DATA, 
    x="Surface", 
    y="PathLength_Total",
    hue="Vision", 
    dodge=True, 
    jitter=0.2, 
    ax=ax
)
ax.set_ylabel("Total Path Length (cm)")
sns.despine()
plt.show()
```
:::

---

## Step 3: Layering Jittered Points, Boxplots, and Custom Palettes

To better summarize central tendency and spread, we overlay boxplots beneath jitter-dodged raw data points and apply a high-contrast, colorblind-friendly palette.

::: {.panel-tabset}
## R (`ggplot2`)

```{r}
#| label: script02-r-plot3
#| fig-width: 7
#| fig-height: 4.5

cbPalette <- c("#1E6654", "#CC5500", "#954535", "#5B7C99",
               "#8366BD", "#D1C071", "#F0EAD6", "grey20")

# Plot 3: Boxplots with jitter-dodged points
ggplot(data = DATA, aes(x = Surface, y = PathLength_Total)) +
  geom_point(aes(shape = Vision, col = Vision),
             position = position_jitterdodge(dodge.width = 0.5, jitter.width = 0.2),
             alpha = 0.5) +
  geom_boxplot(aes(fill = Vision), width = 0.5,
               position = position_dodge(width = 0.5),
               alpha = 0.5, outlier.shape = NA) + 
  scale_color_manual(values = cbPalette[c(1,2)]) + 
  scale_fill_manual(values = cbPalette[c(1,2)]) +
  theme_bw()
```

## Python (`matplotlib` & `seaborn`)

```{python}
#| label: script02-py-plot3
#| fig-width: 7
#| fig-height: 4.5

cbPalette = ["#1E6654", "#CC5500", "#954535", "#5B7C99", "#8366BD", "#D1C071", "#F0EAD6", "#333333"]

# Step 3: Boxplots with jitter-dodged points
fig, ax = plt.subplots(figsize=(7, 4.5))
sns.boxplot(
    data=DATA, x="Surface", y="PathLength_Total", hue="Vision",
    palette=cbPalette[:2], boxprops=dict(alpha=0.5), showfliers=False, ax=ax
)
sns.stripplot(
    data=DATA, x="Surface", y="PathLength_Total", hue="Vision",
    dodge=True, jitter=0.2, palette=cbPalette[:2], alpha=0.5, ax=ax
)

# Deduplicate legend entries
handles, labels = ax.get_legend_handles_labels()
ax.legend(handles[:2], labels[:2], title="Vision")
ax.set_ylabel("Total Path Length (cm)")
sns.despine()
plt.show()
```
:::

---

## Step 4: Subject Aggregation & Visual Polish

Because individual trials introduce nested variance, we aggregate data to subject-level means (`AVE_DATA`). We also refine labels, typography, and legend positioning to polish the figure.

::: {.panel-tabset}
## R (`ggplot2`)

```{r}
#| label: script02-r-plot4
#| fig-width: 8
#| fig-height: 5

# Aggregate trial data to subject-level averages
AVE_DATA <- DATA %>% 
  select(-Trial) %>%
  group_by(Subject, Vision, Surface) %>%
  summarise(
    across(where(is.numeric), ~ mean(.x, na.rm = TRUE)),
    across(!where(is.numeric), dplyr::first),
    .groups = "drop"
  )

# Plot 4: Formatted subject-level summary
ggplot(data = AVE_DATA, aes(x = Surface, y = PathLength_Total)) +
  geom_point(aes(shape = Vision, col = Vision),
             position = position_jitterdodge(dodge.width = 0.5, jitter.width = 0.2),
             alpha = 0.5) +
  geom_boxplot(aes(fill = Vision), width = 0.5,
               position = position_dodge(width = 0.5),
               alpha = 0.5, outlier.shape = NA) + 
  scale_color_manual(values = cbPalette[c(1,2)]) + 
  scale_fill_manual(values = cbPalette[c(1,2)]) +
  theme_bw() +
  labs(
    title = "Postural Sway: Total Path Length by Surface and Vision",
    subtitle = "Mean values aggregated across trials per subject",
    x = "Surface Condition",
    y = "Total Path Length (cm)",
    shape = "Vision Condition", 
    col = "Vision Condition",
    fill = "Vision Condition"
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 11),
    legend.position = "bottom",
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic", color = "gray30")
  )
```

## Python (`matplotlib` & `seaborn`)

```{python}
#| label: script02-py-plot4
#| fig-width: 8
#| fig-height: 5

AVE_DATA = DATA.groupby(["Subject", "Vision", "Surface", "AgeGroup"], as_index=False).mean(numeric_only=True)

# Step 4: Formatted subject-level summary
fig, ax = plt.subplots(figsize=(8, 5))
sns.boxplot(
    data=AVE_DATA, x="Surface", y="PathLength_Total", hue="Vision",
    palette=cbPalette[:2], boxprops=dict(alpha=0.5), showfliers=False, ax=ax
)
sns.stripplot(
    data=AVE_DATA, x="Surface", y="PathLength_Total", hue="Vision",
    dodge=True, jitter=0.2, palette=cbPalette[:2], alpha=0.5, ax=ax
)

handles, labels = ax.get_legend_handles_labels()
ax.legend(handles[:2], labels[:2], title="Vision Condition", loc="upper right")

ax.set_title("Postural Sway: Total Path Length by Surface and Vision\n", fontweight="bold", fontsize=13)
ax.set_xlabel("Surface Condition", fontweight="bold")
ax.set_ylabel("Total Path Length (cm)", fontweight="bold")
sns.despine()
plt.show()
```
:::

---

## Step 5: Visualizing Within-Subject Trajectories

To show individual changes across conditions, we facet by `Surface` and connect subject points from `Open` to `Closed` vision conditions using light gray trajectory lines.

::: {.panel-tabset}
## R (`ggplot2`)

```{r}
#| label: script02-r-plot5
#| fig-width: 8
#| fig-height: 5

# Plot 5: Within-subject lines faceted by Surface
ggplot(data = AVE_DATA, aes(x = Vision, y = PathLength_Total)) +
  geom_line(aes(group = Subject), color = "gray75", alpha = 0.5, linewidth = 0.5,
            position = position_dodge(width = 0.2)) +
  geom_point(aes(group = Subject, shape = Vision, col = Vision), size = 2, alpha = 0.5,
             position = position_dodge(width = 0.2)) +
  geom_boxplot(aes(fill = Vision), width = 0.2, alpha = 0.4, outlier.shape = NA) +
  facet_wrap(~ Surface) + 
  scale_color_manual(values = cbPalette[c(1,2)]) + 
  scale_fill_manual(values = cbPalette[c(1,2)]) +
  theme_bw() +
  labs(
    title = "Postural Sway: Within-Subject Visual Effects",
    subtitle = "Lines connect individual subjects from Open to Closed conditions",
    x = "Vision Condition",
    y = "Total Path Length (cm)"
  ) +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic", color = "gray30"),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 11),
    legend.position = "none",
    strip.text = element_text(size = 12, face = "bold")
  )
```

## Python (`matplotlib` & `seaborn`)

```{python}
#| label: script02-py-plot5
#| fig-width: 8
#| fig-height: 5

# Step 5: Within-subject trajectory lines faceted by Surface
vis_palette = {"Open": cbPalette[0], "Closed": cbPalette[1]}
g = sns.FacetGrid(AVE_DATA, col="Surface", height=4.5, aspect=1.1)

for surf_name, ax in g.axes_dict.items():
    sub = AVE_DATA[AVE_DATA["Surface"] == surf_name]
    
    # Layer 1: Subject lines
    sns.lineplot(
        data=sub, x="Vision", y="PathLength_Total", units="Subject", 
        estimator=None, color="gray", alpha=0.5, ax=ax
    )
    
    # Layer 2: Overlay scatter points
    sns.scatterplot(
        data=sub, x="Vision", y="PathLength_Total", hue="Vision", 
        style="Vision", palette=vis_palette, s=60, alpha=0.8, legend=False, ax=ax
    )
    
    # Layer 3: Boxplot
    sns.boxplot(
        data=sub, x="Vision", y="PathLength_Total", 
        palette=vis_palette, boxprops=dict(alpha=0.4), showfliers=False, ax=ax
    )
    
    ax.set_title(f"Surface: {surf_name}", fontweight="bold")
    ax.set_xlabel("Vision Condition", fontweight="bold")
    ax.set_ylabel("Total Path Length (cm)", fontweight="bold")

g.fig.subplots_adjust(top=0.8)
g.fig.suptitle("Postural Sway: Within-Subject Visual Effects", fontweight="bold", fontsize=14)
sns.despine()
plt.show()
```
:::

---

## Step 6: Incorporating Between-Subject Factors (`AgeGroup`)

Next, we evaluate how age group interacts with balance conditions. We plot the combined interaction (`Surface:Vision`) along the x-axis and facet by `AgeGroup`.

::: {.panel-tabset}
## R (`ggplot2`)

```{r}
#| label: script02-r-plot6
#| fig-width: 9
#| fig-height: 5

# Create interaction factor for clean aesthetic mapping
AVE_DATA <- AVE_DATA %>% 
  mutate(Condition = interaction(Surface, Vision, sep = ":"))

# Plot 6: Interaction conditions faceted by AgeGroup
ggplot(data = AVE_DATA, aes(x = Condition, y = PathLength_Total)) +
  geom_line(aes(group = Subject), color = "gray75", alpha = 0.5, linewidth = 0.5,
            position = position_dodge(width = 0.2)) +
  geom_point(aes(shape = Condition, col = Condition, group = Subject), 
             size = 2, alpha = 0.5, position = position_dodge(width = 0.2)) +
  geom_boxplot(aes(fill = Condition), width = 0.2, alpha = 0.4, outlier.shape = NA) +
  facet_wrap(~ AgeGroup) + 
  scale_color_manual(values = cbPalette) + 
  scale_fill_manual(values = cbPalette) +
  theme_bw() +
  labs(
    title = "Postural Sway: Between- and Within-Subject Effects",
    subtitle = "Lines connect individual subjects across conditions",
    x = NULL,
    y = "Total Path Length (cm)",
    shape = "Condition", col = "Condition", fill = "Condition"
  ) +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic", color = "gray30"),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 11),
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold")
  )
```

## Python (`matplotlib` & `seaborn`)

```{python}
#| label: script02-py-plot6
#| fig-width: 9
#| fig-height: 5

AVE_DATA["Condition"] = AVE_DATA["Surface"].astype(str) + ":" + AVE_DATA["Vision"].astype(str)
cond_order = ["Firm:Open", "Firm:Closed", "Foam:Open", "Foam:Closed"]
cond_palette = {cond: cbPalette[i] for i, cond in enumerate(cond_order)}

g = sns.FacetGrid(AVE_DATA, col="AgeGroup", height=4.5, aspect=1.2)

for age_group, ax in g.axes_dict.items():
    sub = AVE_DATA[AVE_DATA["AgeGroup"] == age_group]
    
    # Layer 1: Subject lines
    sns.lineplot(
        data=sub, x="Condition", y="PathLength_Total", units="Subject", 
        estimator=None, color="gray", alpha=0.5, ax=ax
    )
    # Layer 2: Overlay scatter points
    sns.scatterplot(
        data=sub, x="Condition", y="PathLength_Total", hue="Condition", 
        style="Condition", palette=cond_palette, s=60, alpha=0.8, legend=False, ax=ax
    )
    # Layer 3: Boxplot
    sns.boxplot(
        data=sub, x="Condition", y="PathLength_Total", order=cond_order,
        palette=cond_palette, boxprops=dict(alpha=0.4), showfliers=False, ax=ax
    )
    
    ax.set_title(f"AgeGroup: {age_group}", fontweight="bold")
    ax.set_xlabel("")
    ax.set_ylabel("Total Path Length (cm)", fontweight="bold")

g.fig.subplots_adjust(top=0.8)
g.fig.suptitle("Postural Sway: Between- and Within-Subject Effects", fontweight="bold", fontsize=14)
sns.despine()
plt.show()
```
:::

---

## Step 7: Multivariate Correlation Matrix

To examine multicollinearity among extracted postural metrics across age groups, we build a scatterplot matrix containing bivariate scatter plots (lower triangle), density distributions (diagonal), and correlation coefficients (upper triangle).

::: {.panel-tabset}
## R (`GGally`)

```{r}
#| label: script02-r-plot7
#| fig-width: 8
#| fig-height: 7

plot_data <- DATA %>% 
  select(AgeGroup, PathLength_Total, PathLength_ML, PathLength_AP, EllipseArea95) 

# Plot 7: Correlogram matrix
ggpairs(
  data = plot_data,
  columns = 2:5,
  mapping = aes(color = AgeGroup, fill = AgeGroup),
  lower = list(continuous = wrap("points", alpha = 0.3, size = 1.2)),
  diag = list(continuous = wrap("densityDiag", alpha = 0.3)),
  upper = list(continuous = wrap("cor", size = 4))
) +
  theme_minimal() +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  theme(
    panel.grid.major = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

## Python (`seaborn`)

```{python}
#| label: script02-py-plot7
#| fig-width: 8
#| fig-height: 7

import seaborn as sns
import matplotlib.pyplot as plt

plot_vars = ["PathLength_Total", "PathLength_ML", "PathLength_AP", "EllipseArea95", "AgeGroup"]
plot_data = DATA[plot_vars]

# Plot 7: Seaborn pairplot matrix
g = sns.pairplot(
    plot_data,
    vars=["PathLength_Total", "PathLength_ML", "PathLength_AP", "EllipseArea95"],
    hue="AgeGroup",
    palette={"Young": cbPalette[0], "Old": cbPalette[1]},
    diag_kind="kde",
    plot_kws={"alpha": 0.4, "s": 20}
)
g.fig.suptitle("Multivariate Correlation Matrix by Age Group", y=1.02, fontweight="bold")
plt.show()
```
:::

---

## Step 8: Multi-Panel Publication Composite Layout

Finally, we arrange distinct figure types into a multi-panel layout:
* **Panel A**: Age distribution histogram.
* **Panel B**: Directional path length scatter plot ($\text{PathLength}_{\text{ML}}$ vs $\text{PathLength}_{\text{AP}}$) with confidence ellipses.
* **Panel C**: Total path length density distributions by condition faceted across age groups.

::: {.panel-tabset}
## R (`patchwork`)

```{r}
#| label: script02-r-plot8
#| fig-width: 8
#| fig-height: 8

# Panel A: Age Histogram
FigA <- ggplot(data = AVE_DATA, aes(x = Age)) +
  geom_histogram(fill = "grey", col = "black", binwidth = 2) +
  scale_x_continuous(limits = c(15, 90), breaks = seq(10, 90, 10)) +
  theme_bw() +
  labs(x = "Age (years)", y = "Count") +
  theme(axis.title = element_text(size = 11, face = "bold"))

# Panel B: ML vs AP Path Length Scatter
FigB <- ggplot(data = AVE_DATA, aes(x = PathLength_ML, y = PathLength_AP)) +
  stat_ellipse(aes(fill = Condition), geom = "polygon", alpha = 0.4, col = "black") +
  geom_point(aes(fill = Condition), col = "black", shape = 21, alpha = 0.6) +
  scale_fill_manual(values = cbPalette) +
  theme_bw() +
  labs(x = "Path Length M/L (cm)", y = "Path Length A/P (cm)") +
  theme(axis.title = element_text(size = 11, face = "bold"), legend.position = "none")

# Panel C: Density Distributions by Condition
FigC <- ggplot(data = AVE_DATA, aes(x = PathLength_Total)) +
  geom_density(aes(fill = Condition), alpha = 0.4, col = "black") +
  scale_fill_manual(values = cbPalette) +
  facet_wrap(~ AgeGroup) +
  theme_bw() +
  labs(x = "Total Path Length (cm)", y = "Density") +
  theme(
    axis.title = element_text(size = 11, face = "bold"),
    legend.position = "bottom",
    strip.text = element_text(size = 11, face = "bold")
  )

# Combine Panels via Patchwork
((FigA + FigB) / FigC) +
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(face = 'bold', size = 14))
```

## Python (`matplotlib` & `seaborn`)

```{python}
#| label: script02-py-plot8
#| fig-width: 8
#| fig-height: 8

import matplotlib.pyplot as plt
import seaborn as sns

fig = plt.figure(figsize=(8, 8))
gs = fig.add_gridspec(2, 2, height_ratios=[1, 1])

# Top Row: Panels A and B
ax_a = fig.add_subplot(gs[0, 0])
ax_b = fig.add_subplot(gs[0, 1])

# Bottom Row: Panel C split by AgeGroup (Young vs Old)
ax_c1 = fig.add_subplot(gs[1, 0])
ax_c2 = fig.add_subplot(gs[1, 1])

cond_colors = {
    "Firm:Open": cbPalette[0], "Firm:Closed": cbPalette[1],
    "Foam:Open": cbPalette[2], "Foam:Closed": cbPalette[3]
}

# Panel A: Age Histogram
sns.histplot(data=AVE_DATA, x="Age", binwidth=2, color="gray", edgecolor="black", ax=ax_a)
ax_a.set_xlim(15, 90)
ax_a.set_xlabel("Age (years)", fontweight="bold")
ax_a.set_ylabel("Count", fontweight="bold")
ax_a.set_title("A", loc="left", fontweight="bold", fontsize=14)

# Panel B: ML vs AP Path Length Scatter
sns.scatterplot(
    data=AVE_DATA, x="PathLength_ML", y="PathLength_AP",
    hue="Condition", palette=cond_colors, alpha=0.6, edgecolor="black", ax=ax_b
)
ax_b.set_xlabel("Path Length M/L (cm)", fontweight="bold")
ax_b.set_ylabel("Path Length A/P (cm)", fontweight="bold")
ax_b.get_legend().remove()
ax_b.set_title("B", loc="left", fontweight="bold", fontsize=14)

# Panel C1: Density Distribution - Young Adults
sns.kdeplot(
    data=AVE_DATA[AVE_DATA["AgeGroup"] == "Young"], x="PathLength_Total", hue="Condition",
    palette=cond_colors, fill=True, common_norm=False, alpha=0.4, ax=ax_c1, legend=False
)
ax_c1.set_xlabel("Total Path Length (cm)", fontweight="bold")
ax_c1.set_ylabel("Density", fontweight="bold")
ax_c1.set_title("Young", fontweight="bold", fontsize=11)
# Add "C" tag to the left density subplot
ax_c1.text(-0.18, 1.05, "C", transform=ax_c1.transAxes, fontweight="bold", fontsize=14)

# Panel C2: Density Distribution - Older Adults
sns.kdeplot(
    data=AVE_DATA[AVE_DATA["AgeGroup"] == "Old"], x="PathLength_Total", hue="Condition",
    palette=cond_colors, fill=True, common_norm=False, alpha=0.4, ax=ax_c2
)
ax_c2.set_xlabel("Total Path Length (cm)", fontweight="bold")
ax_c2.set_ylabel("Density", fontweight="bold")
ax_c2.set_title("Old", fontweight="bold", fontsize=11)

# Extract legend handles from ax_c2 and create a clean shared figure legend
handles, labels = ax_c2.get_legend_handles_labels()
ax_c2.get_legend().remove()

fig.legend(
    handles, labels, loc="lower center", 
    bbox_to_anchor=(0.5, -0.01), ncol=4, frameon=False
)

sns.despine()
plt.subplots_adjust(bottom=0.12, hspace=0.35, wspace=0.28)
plt.show()
```
:::
